How to write a PhD in Biological Sciences: a guide for the uninitiated

### 2023-12-28 1.02-useR.Rmd
Add files via upload
# updates to several chapters

Some 30 odd years ago, as an undergraduate, I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the 'carriage-return' button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R (@rcoreteam2021r; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.

>>>>

Some 30 odd years ago, as an undergraduate, I was told that I would have to use computers in one of my 2nd year courses. The thought filled me with dread and I remember putting off going into the computer room for as long as I could. It was a steep learning curve, and I struggled. As my only previous experience had been with a typewriter, each time the cursor got to the end of the screen I hit the 'carriage-return' button. And each machine had an odd white device attached called a mouse, and I had no idea what that did. All this seems quite preposterous now, but imagine embarking on a PhD in biological sciences without reconciling yourself to using a PC. This is practically where we are now with R (@rcoreteam2021language; Figure \@ref(fig:Rlogo)), and a wealth of other emerging and alternative programming platforms.

### 2023-12-28 2.09-databases.Rmd
Add files via upload
# updates to several chapters

Databases of scientific literature have been around since the 1960s, when the number of journals and papers started increasing so much that scholars simply couldn’t keep up. Quite simply, you should use one or more of these databases to look for relevant literature to read. The first database, and one that continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used. As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore, a highly effective way of finding related and pertinent literature in your field.

@@

@@

@@

## Google Scholar, Web of Science or Scopus?

@@

Well, it seems that there are some people who have already thought about this, and come up with a good idea of exactly what's different. Martín-Martín et al. [-@martin-martin2018google] have done a great job of analysing some 2.5 million citations. What they found inspired me to write this section, in which I've focussed on the results from the Life-Sciences. But I encourage you to go read the article for yourself.

I have been known to take the odd peek at my Google Scholar profile over the year, and see how the citations are accumulating. I rarely check on WoS or Scopus, because it's a bit of a faff getting signed in and doing the search. Plus it looks so much smaller when one is habituated to seeing those double digits in GS! However, I've always been a bit uneasy about citing my GS citation rate, H-index or i10 (among other mertics that they give) as I've never really known what all that extra represents. Something grey and unseemly? Well, it turns out that it's all good stuff, and perhaps GS is the better one to cite as it's a more inclusive index: more inclusive of different document types and different languages.

@@

To directly compare your citation metrics from different platforms, and also to get more unbiased metrics than these platforms provide, consider using the software Publish or Perish [@harzing2007publish]. This also provides a great platform on which to harvest data from Google Scholar into a database, or put it directly into another database, like EndNote, Zotero or Mendeley.

## What’s the difference between indices?

According to Martín-Martín et al. [-@martin-martin2018google], the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus. If you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world. Google Scholar results reflect a more inclusive citation index [@martin-martin2018google]. While WoS and Scopus aren't exclusively English or journal publications, they are mostly. But that extra third that GS gives you allows you to show the extra scope that your work is getting outside that English journal mainstream. Is your GS score more than a third higher than your WoS or Scopus score? If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that.

>>>>

Databases of scientific literature have been around since the 1950s, when the number of journals and papers started increasing so much that scholars simply couldn’t keep up [@garfield1955citation]. The first database (from 1962), and one that continues to be one of the largest, was the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) that sought to record not only the contents of selected journals (titles, abstracts, authors and other metadata), but also the citations that each paper used.

Many more literature databases have emerged over the years. Perhaps most notably [PubMed](https://pubmed.ncbi.nlm.nih.gov/) contains a database of biomedical literature and has been free for public use since 1996 and is funded by the US National Institutes of Health ([NIH](https://www.nih.gov/)). The Science Citation Index was acquired by the Thompson Corporation in 1992, and this scholarly information part of the business was sold to private for-profit owners, Clarivate Analytics, in 2016. A rival literature database, [Scopus](https://www.scopus.com/) was founded by publishing company Elsevier in 2004, claiming to be the largest citation database [@schotten2017brief]. Similarly, [Dimensions](https://www.dimensions.ai/), was launched in 2018 as a project owned by Holtzbrinck Publishing Group, owners of Springer-Nature [@vannoorden2018science; @hook2018dimensions]. These initiatives of major science publishing houses are thought to be linked to their attempts to capture the scientific work-flow [@brembs2021replacing]: and see [Part I](#capture).Google launched [Google Scholar](http://scholar.google.co.za/) in 2004, and while this database is free to use, the owners are a for-profit company and the data behind the database remains frustratingly closed.

In 2022, a new Open and free database, [OpenAlex](https://openalex.org/about), successfully aggregated data from [ORCID](https://orcid.org/), [ROR](https://ror.org/), [DOAJ](https://doaj.org/), [Unpaywall](https://unpaywall.org/), [Pubmed](https://pubmed.ncbi.nlm.nih.gov/), [Pubmed Central](https://www.ncbi.nlm.nih.gov/pmc/) and [the ISSN International Centre](https://www.issn.org/) among [others](https://docs.openalex.org/api-entities/sources/get-lists-of-sources) [@singhchawla2022massive]. [OpenAlex](https://openalex.org/) is remarkable for two reasons: (1) by aggregating works from so many different Open Source databases, it has amassed [243 million works](https://openalex.org/about) - nearly 3 times that of Scopus or Web of Science collections. (2) OpenAlex is a free index (specifically it is [CC0](https://creativecommons.org/public-domain/cc0)), it is [Open Source](https://github.com/orgs/ourresearch/repositories?language=&q=openalex&sort=&type=public) and the data is [free to use](https://docs.openalex.org/additional-help/faq#how-is-openalex-licensed). Contrast this with the other for-profit databases, and you will understand that we should support this new initiative.

As a postgraduate student, you should use one or more of these databases to look for relevant literature to read. As we have seen in the last chapter, citations record the way in which people find the literature useful, and are therefore, a highly effective way of finding related and pertinent literature in your field.

@@

**Struggling to define your own keywords?** Help is at hand with a new workshop designed specifically to enable you to determine your own keywords specific to your study, and set up a citation alert from your favourite database (click [here](https://johnmeasey.github.io/Thesis-Writing-Guide/keywords.html)).

@@

## For-profit databases vs Open Source

After the purchase of the Science Citation Index (now renamed [Web of Science](https://www.webofknowledge.com/)) in

@@

## Google Scholar, Web of Science, Dimensions, Scopus or OpenAlex?

@@

According to Martín-Martín et al. [-@martin-martin2018google], who analysed some 2.5 million citations, the GS score is likely to be inflated - 1.90 for GS/WoS and 1.45 for GS/Scopus. If you deviate from this with a higher score, you can give yourself a pat on the back for having work that's reaching more people in more parts of the world. Google Scholar results reflect a more inclusive citation index [@martin-martin2018google]. While WoS and Scopus aren't exclusively English or journal publications, they are mostly. But that extra third that GS gives you allows you to show the extra scope that your work is getting outside that English journal mainstream. Is your GS score more than a third higher than your WoS or Scopus score? If yes, then your work is having a greater impact elsewhere in the world, and there's nothing wrong with that. Perhaps GS is the better one to cite as it's a more inclusive index: more inclusive of different document types and different languages.

@@

To directly compare your citation metrics from different platforms, and also to get more unbiased metrics than these platforms provide, consider using the software Publish or Perish [@harzing2007publish]. This also provides a great platform on which to harvest data from Google Scholar into a database, or put it directly into [your own database](#referencesoftware), like EndNote, Zotero or Mendeley.

We anticipate future studies that will elucidate how Google Scholar, Web of Science, and Scopus differ from Dimensions and OpenAlex. We already know that larger databases should have potential for larger citation numbers. Because OpenAlex is Open Source, this means that you have the potential to build your own API to generate any metric that you are interested in completely free of charge. The prestigious French Sorbonne has been the first university to abandon their subscription to for-profit literature databases [@jack2023sorbonne], and it will be interesting to see how OpenAlex will transform the scene of literature databases in the future.

### 2023-12-28 3.02-Abstract.Rmd
Add files via upload
# updates to several chapters

## What is the abstract?
The abstract is a concise paragraph that sums up the major points of your manuscript so that the potential reader will be able to assess whether or not they want to read the entire paper. It is an abstract of the entire document.

@@

None of this is easy, and you should not expect to write your abstract in a single sitting. It will likely require multiple iterations, and some intense word-smithing to make it as good as it can be. Abstracts almost always have a word limit, and that makes it challenging and means that you have to be concise [see part 2](#concise).

@@

@@

>>>>

## What is the abstract?
The abstract is a concise paragraph that sums up the major points of your manuscript so that the potential reader will be able to assess whether or not they want to read the entire paper. It is an abstract of the entire document. The abstract is a standalone text. The reader should not have to consult more widely in order to make sense of what they are reading  - hence there are no citations in an abstract.

@@

None of this is easy, and you should not expect to write your abstract in a single sitting. It will likely require multiple iterations, and some intense wordsmithing to make it as good as it can be. Abstracts almost always have a word limit, and that makes it challenging and means that you have to be concise [see part 2](#concise).

@@

## Can I use citations in the abstract?
No. This is not a good idea as it defeats the object of the abstract as a standalone text. There are some legitimate times when you might need to use a citation in an abstract (although they are very rare). For example, if all of your manuscript refers to a single previously published paper, and the abstract does not make sense without knowledge of this publication. If you really need to, you can read more about how to cite in the abstract in [Part 1](#citations-in-the-abstract).

@@

**Still struggling with your abstract?** Take advantage of a new workshop that specifically aims to help you write the abstract of your own thesis chapter (click [here](https://johnmeasey.github.io/Thesis-Writing-Guide/abstractwriting.html)).

### 2023-12-28 3.03-Introduction.Rmd
Add files via upload
# updates to several chapters

| **Paragraph** | **Description of contents** |
|--------|---------|
| Paragraph 1 | Overview of the major theme. |
| Paragraph 2 | Identify knowledge gaps in the field. |
| Paragraph 3 | Identify the problem and the gaps you intend to fill. Introduce the important variables that your hypothesis includes. It’s not impossible to mention others, just don’t get distracted. |
| Paragraph 4 | Introduce the approach that you are using, and the organism of choice. |
| Paragraph 5 | Clearly state the hypotheses that are to be tested. |

@@

>>>>

| **Paragraph** | **Description of contents** |
|--------|---------|
| Paragraph 1 | Overview of the relevance of the major theme  |
| Paragraph 2 | Identify knowledge gaps in the field. |
| Paragraph 3 | Identify the problem and the gaps you intend to fill. Introduce the important variables that your hypothesis includes. It’s not impossible to mention others, just don’t get distracted. |
| Paragraph 4 | Introduce the approach that you are using, and the organism of choice. |
| Paragraph 5 | Clearly state the hypotheses that are to be tested. |

@@

**Still struggling with your introduction?** Take advantage of a new workshop that specifically aims to help you write the introduction to your own thesis chapter (click [here](https://johnmeasey.github.io/Thesis-Writing-Guide/formula1.html)).

### 2023-12-28 3.10-Figures.Rmd
Add files via upload
# updates to several chapters

# Figures

@@

Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2021r]. The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](#vector). The drawback is that you might need to invest some time to learn how the graphics package works, and how to manipulate it to do what you want.

>>>>

# Figures {#figures}

@@

Note that the principal software used for drawing data in Dougherty & Ilyankou's, Kieran’s and Claus’s books is R ([see part 1](#useR)), and the many packages that have been developed for use within the R environment [@rcoreteam2021language]. The distinct advantages that this produces are the almost limitless additions and details that you can make to these figures to produce exactly what you want, in a [vector format](#vector). The drawback is that you might need to invest some time to learn how the graphics package works, and how to manipulate it to do what you want.
